Communicative pressures at the semantics-pragmatics interface:\\ Learning biases may prevent the lexicalization of pragmatic inferences

### 2016-09-28 paper/paper.tex
minor changes in paper draft

\subsection{Replication \& expressivity}
Communicative efficiency, or expressivity, has received particular attention from investigations using evolutionary game theory \citep{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}. Under this view, a type's success in communication confers it a higher fitness relative to less successful ones. As a consequence they replicate more than other types, increasing their proportion in the population. This association of a type's communicative success within a population with changes in the types present in it creates a feedback loop that pressures the population towards greater expressivity. The replicator equation gives us the means to make these dynamics precise.

The proportion of types in a given population is captured by a vector $x$, where $x_i$ is type $i$'s proportion in the population. The fitness of a type $i$, $f_i$, is given by its expected utility in this population, $f_i = \sum_j x_j \text{EU}(t_i,t_j)$. That is, its fitness is the sum of its expected communicative success with other types weighted by the latter type's population share. The expected utility of $i$ and $j$ is obtained by considering the expected utility of speaker $i$ interacting with hearer $j$, and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$. $U_S(x,y)$ and $U_R(x,y)$ are respectively $\sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')$ and $U_S(y,x)$ for $n$ and $o$ being the reasoning level of $x$ and $y$, and $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.\footnote{Note that the definition of $U_R(\cdot,\cdot)$ implies equal sender and receiver payoff in an interaction. This need not be so in the general case but suffices for our application.} This quantity is symmetric, reflecting the probability of two types' mutual understanding. Lastly, the average fitness of the population is captured by $\Phi$, $\Phi = \sum_i x_i f_i$. This term serves as a normalizing constant for the (discrete) replicator equation; $\dot{x}_i = \frac{x_i f_i}{\Phi}$

Under its biological interpretation, the replicator equation captures the idea of fitness-relative selection whereby fitter types produce more offspring, leading to their propagation in subsequent generations. In analogy to this kind of replication, many aspects of natural language are subject to processees of transmission and change across varied time-spans. For example, the replicator equation can be understood as a learning across generations as e.g. in \citealt{nowak+etal:2002}, but also as a process of horizontal adaptation (see \citealt[\S3.3]{benz+etal:2005b} for discussion). In the following, we take the latter view in assuming that interlocutors adapt their lexica and their use to that which works best within their population. It should be stressed, however, that the model itself is compatible with either view.

Nowak et al. did not only consider replication, but also recognized the important role of the variation that is introduced by a language's transmition across generations, construed as mutation. Due to this process, the offspring of a type may end up adopting a different type than that of its parent. Crucially, in this work mutation rates were modelled as begin independent from a type. This means that the variation introduced by generational turnovers did not depend on factors such as the relative learnability of a type. To address the issue of selecting a particular type over (near) functional equivalents, we turn to a different strand of research in cultural evolution: {\em iterated learning}.

\subsection{Mutation \& learning}\label{sec:mutation}
Iterated learning is a process in which the behavior of one individual serves as learning input for another, who's behavior subsequently serves as input for a new learner, and so on. For linguistic purposes this process can be thought of as chains of parents and children, where the parent produces linguistic data from which the child infers a language. The latter, now a parent, goes on to produce linguistic data for a new generation of na\"ive learners. Following \citet{griffiths+kalish:2007} we model learning as a process of Bayesian inference in which learners combine the likelihood of a type producing the learning data with prior inductive biases. They then select a type to adopt from the resulting posterior distribution.

Due to the pressure towards learnability it exherts, iterated learning generally leads to simpler and more regular languages (surveys of empirical data and models are given in \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016}). Importantly, experimental and mathematical results suggest the results of this process to reflect learners' learning biases, codified in the following as a prior $P \in \Delta(\mathcal{T})$. A way to think about this bias is as the amount of data a learner would require in order to adopt a language -- or, in our case, a combination of a lexicon and a signaling behavior (cf. \citealt[450]{griffiths+kalish:2007}). Crucially, the extent of the prior's influence has been shown to strongly depend on the learning strategy assumed to underly the inference process. While simulation results suggested that weak biases could be magnified by exposing learners to only small data samples \citep{brighton:2002}, the mathematical characterization provided by \citet{griffiths+kalish:2007} showed that, instead, iterated learning converged to the prior. That is, the distribution over languages in a population or, from an individual's perspective, the likelihood of learning a language corresponds to the learners' prior distribution, irrespective of the amount of input given to learners. This divergence in predictions can be traced back to differences in the selection of hypotheses from the posterior. On the one extreme, Griffith \& Kalish's convergence to the prior holds for learners that sample from the posterior. On the other, more deterministic strategies such as the selection of the type with the highest posterior probability, so-called {\it maximum a posterior estimation} (MAP), increase the prior's influence \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we parametrize the posterior, $P(t_i|d)^l$, to obtain a range of learning strategies that live in the range between posterior sampling and MAP, $l \geq 1$. When $l = 1$ learners sample from the posterior. As $l$ increases towards infinity, the learners' tendency maximize the posterior increases.

More generally, we combine the replicator dynamics with iterated learning by codifying the latter as a transition matrix $Q$. Just as in standard mutator dynamics, $Q_{ij}$ indicates the probability of the children of a parent of type $i$ adopting type $j$. However, to make this process depend on a type's learnbility, this quantity is proportional to the probability of $i$ producing the learning data and that of $j$ given the data.

The elements of the set of learning data $D$ are sequences of length $k$ of state-message pairings. That is, a sequence of observations of language use. Put differently, a datum $d \in D$ contains $k$ members of the set $\{\tuple{s_i,m_j} | s_i \in S, m_j \in M\}$ and $D$ is the set of all such sequences. Having fixed $D$,
\[
 Q_{ij} \propto \sum_{d \in D} P(d|t_i) F(t_j,d),
\]
where $F(t_j,d) \propto P(t_j|d)^l$ and $P(t_j|d) \propto P(t_j) P(d|t_j)$. Given a type $i$, $P(d|t_i)$ can be straightforwardly computed based on $t_i$'s production behavior.

\subsection{Summary}
We argued for expressivity, learnability and simplicity as important pressures that apply on the cultural evolution of language. They are respectively modelled as communicative efficiency-relative replication, iterated Bayesian learning, and a prior that biases learners for compressed lexical meanings. Taken together these evolutionary dynamics are described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:

\[ 
\hat{x_i} = \sum_j Q_{ji} \frac{x_jf_j}{\Phi}
\]

The basic units that the dynamics operate on are a combination of lexica lexicon and their use; a type. A type's expressivity depends on its communicative efficiency within a population while its learnability depends on the fidelity by which it is infered by new generations of na\"ive learners.

In sum, the innovation of this model lies in its combination of functional pressure on successful communication, effects of learning biases on (iterated) Bayesian language learning \citealt{griffiths+kalish:2007}, and probabilistic models of language use in populations with distinct lexica \citep{frank+goodman:2012,franke+jaeger:2014,bergen+etal:2016}. In particular, this synthesis enables for the investigation of the effects of communicative pressures on the semantics-pragmatics interface. In doing so, it links the previously disconnected areas of rational probabilistic language use and cultural evoltion.

>>>>

\subsection{Replication \& expressivity}
Expressivity has received particular attention from investigations using evolutionary game theory \citep{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}. Under this view, a type's communicative success confers it a higher fitness, a measure that is relative to the success of other types in the population. In the simplest models fitness directly translates into the proportion of types present in the subsequent population. As a consequence, fitter types flourish while unfit ones are driven out. This association of communicative success within a population with changes in the proportion of types present in it creates a feedback loop that pressures the population towards greater expressivity. The replicator equation gives us the means to make these dynamics precise.

The proportion of types in a given population is codified in a vector $x$, where $x_i$ is type $i$'s proportion. The fitness of a type $i$, $f_i$, is given by its expected utility in this population, $f_i = \sum_j x_j \text{EU}(t_i,t_j)$. That is, its fitness is the sum of its expected communicative success in interacting with other types weighted by the latter's population share. The expected utility of $i$ and $j$ is obtained by considering the expected utility of speaker $i$ interacting with hearer $j$ and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$. $U_S(x,y)$ and $U_R(x,y)$ are $\sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')$ and $U_S(y,x)$, respectively, for $n$ and $o$ being the reasoning level of $x$ and $y$, and $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise. This quantity is symmetric, reflecting the probability of two types' mutual understanding. Lastly, the average fitness of the population is given by $\Phi$, $\Phi = \sum_i x_i f_i$. This term serves as a normalizing constant for the (discrete) replicator equation: $\dot{x}_i = \frac{x_i f_i}{\Phi}$

Under its biological interpretation the replicator equation captures the idea of fitness-relative selection whereby fitter types produce more offspring, leading to their propagation in subsequent generations. Similarly, many aspects of natural language are subject to processees of transmission and change across varied time-spans. Amongst others, the replicator equation can be understood as a model of language acquisition across generations, as e.g. in \citealt{nowak+etal:2002}, but also as a process of horizontal adaptation (see \citealt[\S3.3]{benz+etal:2005b} for discussion). In the following, we take the latter view in assuming that interlocutors may adapt their types to that which works best within their population.

In their series of papers on language evolution, Nowak and colleagues did not only consider replication, but also recognized the central role that variation plays in its transmission. In keeping the analogy to evolutionary processes, this variation can be likened to mutation to the effect that a type's offspring may adopt a different type than that of its parent. However, mutation rates were modeled as independent from a type \hl{CITATION} meaning that the variation introduced by generational turnovers did not depend on factors such as the relative learnability of a type. In the following, we interpret mutation as learnability to address the issue of selecting a particular type over (near) functional equivalents. For this purpose, we turn to a different strand of research in cultural evolution: {\em iterated learning}.

\subsection{Mutation \& learning}\label{sec:mutation}
Iterated learning is a process in which the behavior of one individual serves as learning input for another, who's behavior subsequently serves as input for a new learner, and so on. In language this process can be thought of as a progression through chains of parents and children where the parent produces linguistic data from which the child infers a language. The latter, now a parent, goes on to produce linguistic data for a new generation of na\"ive learners. Following \citet{griffiths+kalish:2007} we model iterated learning as a process of Bayesian inference in which learners combine the likelihood of a type producing the learning data with prior inductive biases. They then select a type to adopt from the resulting posterior distribution.

Due to the pressure towards learnability it exherts, iterated learning generally leads to simpler and more regular languages (see \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016} for recent surveys). Importantly, experimental and mathematical results suggest the outcome of this process to reflect the learners' a priori biases. In a Bayesian setting these biases are codified in the prior $P \in \Delta(\mathcal{T})$. A way to think about this prior is as the amount of data a learner would require in order to adopt a language (cf. \citealt[450]{griffiths+kalish:2007}). Or, in our case, a combination of a lexicon and a signaling behavior. \hl{INTRODUCE SIMPLICITY BIAS HERE}

The extent of the prior's influence has been shown to strongly depend on the learning strategy assumed to underly the inference process. While simulation results suggested that weak biases could be magnified by exposing learners to only small data samples \citep{brighton:2002}, the mathematical characterization provided by \citet{griffiths+kalish:2007} showed that, instead, iterated learning converged to the prior. That is, the distribution over languages in a population, or the likelihood of learning a language when taking an individual's perspective, corresponds to the learners' prior distribution irrespective of the amount of input given to learners. This divergence in predictions can be traced back to differences in the selection of hypotheses from the posterior. On the one extreme, Griffith \& Kalish's prediction of convergence to the prior holds for learners that sample from the posterior. On the other, more deterministic strategies such as the selection of the type with the highest posterior probability, so-called {\it maximum a posterior estimation} (MAP), increase the prior's influence \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we parametrize the posterior, $P(t_i|d)^l$, to obtain a range of learning strategies that live between posterior sampling and MAP, $l \geq 1$. When $l = 1$ learners sample from the posterior. As $l$ increases towards infinity, the learners' propensity to maximize the posterior increases.

We combine the replicator dynamics with iterated learning by codifying the latter as a transition matrix $Q$. Just as in standard mutator dynamics, $Q_{ij}$ indicates the probability of the children of a parent of type $i$ adopting type $j$. However, to make this process depend on a type's learnbility, this quantity is proportional to the probability of $i$ producing the learning data and that of inferring $j$ given the data.

The elements of the set of learning data $D$ are $k$-length sequences of state-message pairings. That is, a sequence of observations of language use. Put differently, a datum $d \in D$ contains $k$ members of the set $\{\tuple{s_i,m_j} | s_i \in S, m_j \in M\}$. $D$ is the set of all such sequences. With the at hand $D$, the learning matrix is constructed as follows:
\[
 Q_{ij} \propto \sum_{d \in D} P(d|t_i) F(t_j,d),
\]
where $F(t_j,d) \propto P(t_j|d)^l$ and $P(t_j|d) \propto P(t_j) P(d|t_j)$. Given a type $i$, $P(d|t_i)$ can be straightforwardly computed based on $t_i$'s production behavior.

\subsection{Summary}
We argued that expressivity, learnability and simplicity are central pressures that apply on the cultural evolution of language. They are respectively modelled as communicative efficiency-relative replication, iterated Bayesian learning, and a prior that biases learners for compressed lexical meanings. Taken together their dynamics can be described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:

\[ 
\hat{x}_i = \sum_j Q_{ji} \frac{x_jf_j}{\Phi}
\]

The units that the dynamics operate on are a combination of a lexicon and linguistic behavior which determines its use, i.e., a type. A type's expressivity depends on its communicative efficiency within a population while its learnability depends on the fidelity by which it is infered by new generations of na\"ive learners.

%In sum, the innovation of this model lies in its combination of functional pressure on successful communication, effects of learning biases on (iterated) Bayesian language learning \citealt{griffiths+kalish:2007}, and probabilistic models of language use in populations with distinct lexica \citep{frank+goodman:2012,franke+jaeger:2014,bergen+etal:2016}. In particular, this synthesis enables for the investigation of the effects of communicative pressures on the semantics-pragmatics interface. In doing so, it links the previously disconnected areas of rational probabilistic language use and cultural evoltion.
